\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-08-03 mers-structure.tex
Minor text changes.

Although such traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
Between the two trees in figure \ref{recombinant_features}B four `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
For the reverse comparison there are five `human' clades in fragment 2 that are expanded.
All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
This model estimates 4 parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every 20 000 steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left 70 000 states on which to base posterior inference.

>>>>

Although such traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014,quick_rapid_2015}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

@@

%%% AR - this paragraph is a bit of a mess. Suggests we can't trust most of the analysis above. If we can't believe the camel tree why do we trust a structured coalescent model?
Conversely, our results strongly suggest that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination will interfere with the inference of human outbreak clusters (Figure \ref{recombinant_features}A).
Between the two trees in figure \ref{recombinant_features}B four (out of 54) `human' clades are expanded where either singleton introductions or two-taxon clades in fragment 2 join other clades in fragment 1.
For the reverse comparison there are five `human' clades (out of 53) in fragment 2 that are expanded.
All such clades have low divergence (figure \ref{recombinant_features}B) and thus incongruences in human clades are more likely to be caused by differences in deme assignment rather than actual recombination.
And while we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (Figure \ref{recombinant_features}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (Figure \ref{flower}).
Critically, we observe the same source-sink dynamics between camel and human populations in trees constructed from separate genomic fragments as were observed in the original full genome tree (Figures \ref{mcc}, \ref{recombinant_features}B).
Observed departures from strictly clonal evolution suggest that while recombination is an issue for inferring MERS-CoV phylogenies, its effect on the human side of MERS outbreaks is minimal, as expected.
MERS-CoV evolution on the reservoir side, though complicated by recombination, is nonetheless still amenable to phylogenetic methods, in part through limited diversity of the virus in camels (see next section).
In humans MERS-CoV evolution should be far easier to track as the only detectable and problematic recombinants are more likely to arise within the transmission chain, than through human co-infection with distinct MERS-CoV lineages.

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
This model estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every 20 000 steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left 70 000 states on which to base posterior inference.
